{\tt APFELgrid}: a fast interface to hadronic observables for fits of parton densities

### 2016-05-01 paper.tex
updating draft

\cortext[author] {Corresponding author.\\\textit{E-mail address:} nathan.hartland@physics.ox.ac.uk}
\address[a]{Rudolf Peierls Centre for Theoretical Physics,\\ 1 Keble Road, University of Oxford, OX1 3NP, Oxford, UK}
\address[b]{Theory Department, CERN, CH-1211 Geneva 23, Switzerland}

@@

The {\tt FastKernel} method, developed in the framework of
the NNPDF global analyses~\cite{Ball:2010de}, was developed to address
this problem. This method differs from the standard procedure \`{a} la {\tt APPLgrid} or {\tt FastNLO} in that it
maximises the amount of information that is precomputed prior to fitting such as to minimise the amount
of operations to be performed during the fit. Specifically, the {\tt FastKernel} method relies on the precomputation
of the partonic hard cross-sections and DGLAP evolution kernels
and their combination into a single look-up table, here called a {\tt FastKernel} ({\tt FK}) table, in such a way that the prediction for a given hadronic
observable can be obtained by a simple matrix product between
the respective {\tt FK} table and the initial scale PDFs.

@@

The {\tt APFELgrid} plugin generates {\tt FK} tables of the form
expressed in Eq.~(\ref{eq:FKTable}) from weight grids supplied by {\tt
  APPLgrid} files. In this section we shall examine the performance
differences between the two expressions for cross-section prediction
Eq.~(\ref{eq:applconv}) ({\tt APPLgrid}) and Eq.~(\ref{eq:FK}) ({\tt
  FK}).
%
For a comprehensive benchmark, we consider here a wide range of
processes including LHC and Tevatron electroweak vector boson
production measurements, $t\bar{t}$ total cross-sections,
double-differential Drell-Yan cross-sections and inclusive jet
data. Predictions are provided for a wide range of kinematics, for a
total of 52 source {\tt APPLgrid} files roughly corresponding to the
full collider dataset of the forthcoming NNPDF3.1 PDF determination.

In Fig.~\ref{fig:timings} we compare the convolution time per point
between {\tt APFELgrid} and {\tt APPLgrid} for all the 52
processes. We show timings for the {\tt APFELgrid} library in four
different configurations: AVX-OpenMP 2x (2 CPUs), AVX SSE3, and
standard double precision dot product convolution. When comparing the
results we observe a great performance improvement for convolution
time per point when using {\tt APFELgrid}. In particular, when
comparing {\tt APFELgrid} AVX to {\tt APPLgrid} timings we obtain a
minimum of x10 improvement for electroweak vector boson production and
a maximum of x2000 for inclusive jet data. These results confirm the
great usefulness of {\tt APFELgrid} in terms of performance for fits
of PDFs.

In Fig.~\ref{fig:performance} we provide information about the {\tt
  APFELgrid} table generation, grid size and memory usage. In terms of
generation time per point, we observe a timings from few milliseconds
to 3.5 minutes per point. Such difference is almost due to the density
of weights contained in the input {\tt APFELgrid} requiring a larger
number of combination operations. When considering the grid size, we
confirm, as expected, that in several processes the {\tt APPLgrid}
files are smaller than {\tt APFELgrid} ones. This feature is justified
by the lack of compression of {\tt APFELgrid} files. Finally, when
measuring the resident set size memory used by the convolution
programs, the amount of memory required by {\tt APFELgrid} is smaller
than {\tt APPLgrid} for all processes considered here.

>>>>

\cortext[author] {Corresponding author.\\\textit{E-mail address:} nathan.hartland@physics.ox.ac.uk}
\address[a]{Rudolf Peierls Centre for Theoretical Physics,\\ 1 Keble Road, University of Oxford, OX1 3NP, Oxford, UK}
\address[b]{Theoretical Physics Department, CERN, Geneva, Switzerland}

@@

The {\tt FastKernel} method, developed in the framework of the NNPDF
global analyses~\cite{Ball:2010de,Ball:2014uwa}, was developed to
address this problem. This method differs from the standard procedure
\`{a} la {\tt APPLgrid} or {\tt FastNLO} in that it maximises the
amount of information that is precomputed prior to fitting such as to
minimise the amount of operations to be performed during the
fit. Specifically, the {\tt FastKernel} method relies on the
precomputation of the partonic hard cross-sections and DGLAP evolution
kernels and their combination into a single look-up table, here called
a {\tt FastKernel} ({\tt FK}) table, in such a way that the prediction
for a given hadronic observable can be obtained by a simple matrix
product between the respective {\tt FK} table and the initial scale
PDFs.

@@

The {\tt APFELgrid} plugin generates {\tt FK} tables of the form
expressed in Eq.~(\ref{eq:FKTable}) from weight grids supplied by {\tt
  APPLgrid} files. In this section we shall examine the performance
differences between the two expressions for cross-section prediction
Eq.~(\ref{eq:applconv}) ({\tt APPLgrid}) and Eq.~(\ref{eq:FK}) ({\tt
  FK}).
%
For a comprehensive benchmark, we consider here a wide range of
processes including LHC and Tevatron electroweak vector boson
production
measurements~\cite{Aaij:2012mda,Aaij:2012vn,Chatrchyan:2013mza,Chatrchyan:2013uja,Chatrchyan:2012xt,Aad:2013iua,Aad:2011fp,Aad:2011dm,Aaltonen:2010zza},
$t\bar{t}$ total
cross-sections~\cite{ATLAS:2012aa,ATLAS:2011xha,TheATLAScollaboration:2013dja,Chatrchyan:2013faa,Chatrchyan:2012bra,Chatrchyan:2012ria},
double-differential Drell-Yan
cross-sections~\cite{Chatrchyan:2013tia,CMS:2014jea} and inclusive jet
data~\cite{Chatrchyan:2012bja,Aad:2011fc,Aad:2013lpa,Abazov:2007jy}. Predictions
are provided for a wide range of kinematics, for a total of 52 source
{\tt APPLgrid} files roughly corresponding to the full collider
dataset of the forthcoming NNPDF3.1 PDF determination.

In Fig.~\ref{fig:timings} we compare the convolution time per point
between {\tt APFELgrid} and {\tt APPLgrid} for all the 52
processes. We show timings for the {\tt APFELgrid} library in four
different configurations: AVX-OpenMP 2x (2 CPUs), AVX SSE3, and
standard double precision dot product convolution. When comparing the
results we observe a great performance improvement for convolution
time per point when using {\tt APFELgrid}. In particular, when
comparing {\tt APFELgrid} AVX to {\tt APPLgrid} timings we obtain a
minimum of 10x improvement for electroweak vector boson production and
a maximum of 2000x for inclusive jet data. These results confirm the
great usefulness of {\tt APFELgrid} in terms of performance for fits
of PDFs.

In Fig.~\ref{fig:performance} we provide information about the {\tt
  APFELgrid} table generation, grid size and memory usage. In terms of
generation time per point, we observe timings from few milliseconds to
3.5 minutes per point. Such difference is almost due to the density of
weights contained in the input {\tt APPLgrid} which requires a larger
number of combination operations. When considering the grid size we
confirm that in several processes the {\tt APPLgrid} files are smaller
than {\tt APFELgrid} ones as expected. This feature is justified by
the lack of compression of {\tt APFELgrid} files. Finally, when
measuring the resident set size memory used by the convolution
programs, the amount of memory required by {\tt APFELgrid} is smaller
than {\tt APPLgrid} for all processes considered here by at least 75\%.
